\ifx\PREAMBLE\undefined
\input{preamble}
\begin{document}
\fi
\newpage
\section{Integer}
\begin{itemize}
\item Interpretation of binary representations
\begin{equation*}
\begin{array}{lll}
B2U_w(\overrightarrow{x})=\sum\limits_{i=0}^{w-1}x_i2^i & UMin_w=0 & UMax_w=2^w-1\\

B2T_w(\overrightarrow{x})=-x_{w-1}2^{w-1}+\sum\limits_{i=0}^{w-2}x_i2^i & TMin_w=-2^{w-1} & TMax_w=2^{w-1}
-1\\
\end{array}
\end{equation*}

\item Transformation between unsigned and signed
\begin{gather*}
T2U_w(x)=\begin{cases}
x & 0 \le x \le TMax_w\\
2^w+x & TMin_w\le x < 0
\end{cases}\\
U2T_w(u)=\begin{cases}
u & 0 \le u \le TMax_w\\
u - 2^w & TMax_w < u \le UMax_w
\end{cases}\\
\end{gather*}

\item Expansion
\begin{equation*}
\begin{array}{ll}
\overrightarrow{u}=[u_{w-1},\dots,u_0]\rightarrow\overrightarrow{u}'=[0,\dots,0,u_{w-1},\dots,u_0] & B2U_w(\overrightarrow{u})=B2U_{w'}(\overrightarrow{u}')\\
\overrightarrow{x}=[x_{w-1},\dots,x_0]\rightarrow\overrightarrow{x}'=[x_{w-1},\dots,x_{w-1},x_{w-1},\dots,x_0] & B2T_w(\overrightarrow{x})=B2T_{w'}(\overrightarrow{x}')\\
\end{array}
\end{equation*}

\item Truncation
\begin{equation*}
\begin{array}{ll}
\overrightarrow{u}=[u_{w-1},\dots,u_0]\rightarrow\overrightarrow{u}'=[u_{k-1},\dots,u_0] & 
B2U_k(\overrightarrow{u}')=B2U_{w}(\overrightarrow{u})\mod 2^k\\
\overrightarrow{x}=[x_{w-1},\dots,x_0]\rightarrow\overrightarrow{x}'=[x_{k-1},\dots,x_0] & 
B2T_k(\overrightarrow{x}')=U2T_{k}\left(B2U_{w}(\overrightarrow{x})\mod 2^k\right)\\
\end{array}
\end{equation*}

\item Addition
\begin{equation*}
\begin{array}{ll}
x+^u_wy=\begin{cases}
x + y 		& 0 \le x+y < 2^w\\
x + y - 2^w & 2^w \le x+y \le 2^{w+1} - 2
\end{cases} & 0\le x,y \le 2^w-1\\
x+^t_wy=\begin{cases}
x + y + 2^w & -2^w \le x+y < -2^{w-1}\\
x + y 			& -2^{w-1} \le x+y \le 2^{w-1} - 1\\
x + y - 2^w & 2^{w-1} \le x+y \le 2^{w}-2
\end{cases} & -2^{w-1} \le x,y \le 2^{w-1} - 1\\
\end{array}
\end{equation*}

\item Negation
\begin{equation*}
-^t_wx=\begin{cases}
TMin_w & x = TMin_w\\
-x & TMin_w < x \le TMax_w
\end{cases}
\end{equation*}

\item Multiplication
\begin{equation*}
\begin{array}{ll}
x*^u_wy = (x\cdot y)\mod 2^w & u\ll k = u * ^u_w2^k\\
x*^t_wy = U2T_w((x\cdot y)\mod 2^w) & x\ll k = x * ^t_w2^k\\
\end{array}
\end{equation*}

\item Division
\begin{equation*}
\begin{array}{ll}
\begin{gathered}
x\gg k = \lfloor x/2^k \rfloor\\
(x + (1\ll k) - 1)\gg k = \lceil x/2^k \rceil\\
\end{gathered} & 
x/y\equiv\begin{cases}
\lfloor x/y \rfloor & x \ge 0, y>0\\
\lceil x/y \rceil & x < 0, y > 0\\
\end{cases}
\end{array}
\end{equation*}
\end{itemize}
\newpage
\section{Floating Point Number}
\begin{itemize}
\item Definition $V=(-1)^s\times M \times 2^E$. 1-bit $s$ encodes the sign $s$, $k$-bit \texttt{exp} encodes the exponent $E$, and $n$-bit \texttt{frac} encodes the significand $M$. float: 1, 8, 23; double: 1, 11, 52.
 \begin{description}
 	\item[Normalized] \texttt{exp} is neither 0 or $2^k-1$ (all 1). 
 	\begin{itemize}
 		\item $E=e-Bias$, in which $Bias=2^{k-1}-1$. 
 		\item $M=1.f_{n-1}\dots f_1f_0$.
 	\end{itemize} 
 	\item[Denormalized] \texttt{exp = 0}
 	\begin{itemize}
 		\item $E=1-Bias$.
 		\item $M=0.f_{n-1}\dots f_1f_0$.
 	\end{itemize}
 	\item[Infinity] \texttt{exp}=$2^{k}-1$ (all 1), \texttt{frac} = 0.
 	\item[NaN] \texttt{exp}=$2^{k}-1$ (all 1), \texttt{frac} $\neq$ 0.
 \end{description}

\item Use round to even.
\item $x+^fy\equiv Round(x+y)$. FP addition lacks associativity but features monotonicity: $x+a\ge x+b$ if $a\ge b$.
\item $x*^fy\equiv Round(x\times y)$. FP Multiplication is not associative, and it does not distribute over addition. It features monotonicity:
\begin{align*}
a\ge b, c\ge 0 &\Rightarrow a * ^fc \ge b * ^fc\\
a\ge b, c\le 0 &\Rightarrow a * ^fc \le b * ^fc\\
a\neq NaN &\Rightarrow a * ^fa\ge 0
\end{align*}
\item Type conversions:
\begin{itemize}
	\item \texttt{int} to \texttt{float}: no overflow. Possible to be rounded.
	\item \texttt{int/float} to \texttt{double}: precise conversion.
	\item \texttt{double} to \texttt{float}: possible to overflow to infinity. Possible to be rounded.
	\item \texttt{float/double} to \texttt{int}: round to 0. Possible to overflow.
\end{itemize}
\end{itemize}
\ifx\PREAMBLE\undefined
\end{document}
\fi